Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-10 paper.tex
update

In Fig.~\ref{fig:radial_emis} we show the radius with the maximum
luminosity for substructures and smooth density profiles, both for
three different mass scales. We find that the smooth luminosity per
logaritmic radial bin peaks at $x \dd^2 L/\dd x^2+\dd L/\dd x == 0
\rightarrow r \propto \rvir/3c$. Hence the main contribition to the
luminosity for large mass scales, which are usually assosiated with a
low concentrated DM halos, is focused to the regime around 10\% of
$\rvir$.  It clearly shows how that the main contribution from the
substructures is coming from outer parts of the DM halos, while the
main contribution from the smooth density profiles peaks around the
scale radius. Not that, even though the bulk of substructures have
been erased in the central regions of DM halos, a cluster in
projection has a significant enhancement due to substructures at a
radius of just a few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Luminosity dominated radius for different mass
    scales. The solid lines show the emissivity from a cuspy
    $\beta=1.0$ NFW density profile for three different mass scales;
    an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the emissivity from substructures for the same three mass
    scales. All emissivities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$.}
  \label{fig:radial_emis}
\end{figure}

@@

There is also a possiblity to search for dark matter annihilation
indirectly in antimatter channels, like antiprotons or positrons. Here
diffusion is, however, very efficient which means that the directional
signature which is so important for gamma-rays, is completely
erased. This method has, however, recently been much in focus due to
the surprising rise with energy of the electron to proton ratio
measured by the PAMELA satellite experiment \cite{Adriani:2008zr} and
the ``bump'' in the total electron plus positron spectrum detected by
Fermi \cite{Abdo:2009zk}. The viable particle physics models
describing these results contain some non-trivial, or even unnatural
elements, however. First of all, annhilation to quarks has to be very
much suppressed, due to the non-observation of an enhanced antiproton
flux by PAMELA \cite{Adriani:2010rc}. This means that only so-called
{\em leptophilic} models, where the annihilation only goes to leptons
have to be considered. Here, one has to have a sizeable fraction going
to muons and tau leptons, as the shape of the spectrum disfavours
direct annihilation into electrons and positrons. To get agreement
with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult
to achieve with substrucutre alone. An additional enhancement of the
annihilation rate has to be invoked, for example given by the {\em
  Sommerfeld effect}. This basically means that the usual connection
between annihilation rate at decoupling (when relative velocities $v$
were of order $0.2 - 0.3$), which gives the relic density, and
present-day annihilation in the halo (with velocities $v\sim 10^{-3}$)
is broken. This would come about because the dark matter sector
contains additional light particles which mediates an interaction with
almost bound states which enhance the wave function at the relative
origin, and therefore the annihilation rate (for a pedagogical review
of viable scenarios, see \cite{ArkaniHamed:2008qn}). As a final
difficulty of these models, the lack of an inverse-compton signature
from the galactic center means that only shallow density profiles can
describe the galactic dark matter density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the dark matter modelling of these effects, see
\cite{Bergstrom:2009ib}. For a recent treatment, showing still viable
models, see \cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, dark matter would
more likely be explained by the standard scenario of a WIMP, for
example the thoroughly studied lightest supersymmertric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may gave interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical dark matter community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
dark matter halo substructure has been realised for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieive the boost factor of the order of a few
hundred to a thousand in the solar neighbourhood, as substructure
according to numerical simulations is most important in the outer
portions of the halo, due to tidal stripping of subhaloes in the inner
part. Another potentially very important effect which may explain the
large boost had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld may years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of dark matter
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probablity for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}|,
$$ with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the
realtive velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small
velocities. For a Yukawa-like particle of mass $m_\phi$, mediating a
weak attractive force with coupling contant $\alpha_Y$ between dark
matter particles $\chi$ with mass $m_\chi$, the small-velocity limit
of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the dark matter sector, one has to solve
numerically a coupled system of differential equations using
appropriate boundary conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depnding however on the exact parameters of
the theory. The effect is usually strongly velocity-dependent,
depending on velicity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such a a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

@@

\subsubsection{Supersymetric dark matter}
The most studied models for dark matter are supersymmetric ones,
especially models where the lightest supersymmetric particle, in most
models the lightest neutralino, is stable. The stability is assured in
viable SUSY models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for dark matter. In
addition, as the coupling to ordinary matter is given by gauge
couplings, the neutralino is automatically a WIMP
candidate. Unfortunately, the breaking of supersymmetry (which has to
be there since otherwise, e.g., there would exist a scalar electron
with the same mass as the electron) means that a large number of
essentially free parameters enter actual calculations.  The neutralino
is then a linear combination of the supersymmetric partners of the
U(1) gauge boson, one of the SU(2) gauge bosons, and the two Higgs
doublets in the Minimal Supersymmetric extension of the Standard Model
(the MSSM) with mixing coefficients which depend on the SUSY breaking
parameters \cite{1984NuPhB.238..453E}. The parameter space is very
large, too large in general to scan efficiently even with the most
powerful computers. Therefore, often simplifying assumptions are made
to limit the number of free parameters. Even then, a large number of
viable models is found which give a relic density consistent with the
WMAP data. This means that they are indeed very good templates for
WIMP dark matter, something that explains their popularity besides the
subjective statement that few more plausible models for dark matter
have been put forward so far.

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.photon.eps}
\caption{\it Spatial dependence of light from stars and dust. We show
  2D surface brightness profiles in the $r$-band ($\sim 1\,\ev$)
  obtained from stacked clusters in the Sloan Digital Sky Survey
  (SDSS) at the redshift $z \sim 0.25$ \cite{2005MNRAS.358..949Z}. The
  brightness profiles have been weighted with the $\rvir$ normalized
  area inside radius $r$, and trace the radial dependence of the
  luminosity from stars and galaxies. The crosses show the total
  observed starlight including the diffuse intracluster (ICL), the
  galaxies, and the brightest cluster galaxy (BCG) in the center of
  the cluster. The diamonds denote the observed starlight from the
  ICL and the BCG. The solid line show the fit to the data of the
  total light including the ICL, the BCG, and the galaxies, while the
  smooth component is represented by the dashed line that is fitted to
  only the ICL and the BCG.}
 \label{fig:SD_spatial}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eq, where we use have used the mass
  $\mvir=4.0\times10^{13}\msun$ in this figure. Finally we renormalize
  the SD spectra to the radius $r=0.071\rvir$, where the smooth energy
  density of the SD light (see Fig.~\ref{fig:SD_Edens}) equals the
  energy density of the CMB. The red dashed lines show the fitted SD
  spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. The SD light has been renormalized
   to a cluster with the mass $\mvir=4.0\times10^{13}\msun$. Finally we
   show the energy density two magnetic field models with a central
   magnetic field of 3~$\mu G$. The magnetic field scales with the gas
   density to the power $\alpha_\rmn{B}$; green dash-dotted line
   ($\alpha_\rmn{B}=0.5$) and orange dash-dotted line
   ($\alpha_\rmn{B}=0.7$). Note that the SD radiation is dominating
   the energy density inside $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by well
studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the angular extended Virgo cluster. It is quite striking how
constraining the upper limits for the Fornax cluster is compared to
the other clusters, especially in the $1-10$~GeV energy regime where
Fermi-LAT is most sensitive due to the maximized effective area. The
upper limits for Virgo and Perseus are background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. One should also note that the high CRp
induced flux from both Coma and Perseus is of the same magnitude as
the DM flux from LP models, making it difficult for indirect DM
searches in those clusters. Conversely is the Fornax cluster a great
target for indirect DM studies because the relative low gamma-ray flux
from CRps, absents of an active AGN, and high DM gamma-ray
flux. Actually, the Fermi-LAT 3 year data will be able to constrain
the smallest mass of substructures using the DM BM models. We have
also learned that the closest clusters are superior for DM searches
compared to further away clusters, even though the close by clusters
have a mass that is smaller. In fact, we can understand this by
looking at the luminosity-to-mass scaling relations. The gamma-ray
luminosity from the smooth density profile is given by
\begin{equation}
L_{\gamma,\rmn{sm}} \propto \int \dd V \rho(r)^2 \propto \frac{c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity including boost factors for the LP and
the BM models are given by
\begin{eqnarray}
\label{eq:LP_scaling}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \mvir^{0.73}\qquad
\,\,\,\,\,\,\rmn{and}\\
\label{eq:BM_scaling}
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \mvir^{1.06}\,,\qquad
\rmn{where} \\
B_\rmn{LP}&=&B_\sfe B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
B_\rmn{BM}&=&B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the mass with a factor two doubles the luminosity,
but increasing the distance by a factor two decrease the flux with a
factor four. It is also usefull to relate the luminosities to a more
model independent variable as the maximum circular velocity given by
\begin{eqnarray}
V_\rmn{max} &=& V_\rmn{circ}(r_\rmn{max})\qquad\,\,\,\,\,\,\,\,\,\rmn{where}\nonumber\\
V_\rmn{circ}(r)&=& \sqrt{\frac{G_\rmn{N}\,M(<r)}{r}}\qquad\rmn{and}\nonumber\\
\frac{\dd V_\rmn{circ}(r)}{\dd r} &=& 0\quad \Longrightarrow\quad r_\rmn{max}\,.
\end{eqnarray}
We find that $V_\rmn{max}=650\,\rmn{km/s}$ of a large $10^{15}\,\msun$
cluster, and $V_\rmn{max}=420\,\rmn{km/s}$ of a small
$4\times10^{13}\,\msun$ group.

@@

NEED TO REWRITE, JUMPING BACK AND FORTH BETWEEN PANELS In order to
learn about the spatial distribution of the DM flux in various
clusters for different DM models and associated gamma-ray components,
we show in the left panel of Fig.~\ref{fig:SB_clu_nosub} the
brightness profiles for the smooth DM distribution for the same four
representative clusters that was shown previously in
Fig.~\ref{fig:SB_clu}. In addition, we show the spatial dependence of
the individual gamma-ray components for the Fornax cluster in the
right panel. We find that the spatial distribution of the DM emission
at 1~GeV in the outer parts of the all clusters have the same
shape. The shape trace the underlying density slope of $-3$, which is
expected for IC upscattered CMB photons. This picture is confirmed by
the components study in the right panel that show the neglible
contribution from the IC upscattered light from SD at this
energy. Since the surface brightness from BM models are dominated by
the continuum emission at each radial bin in all clusters, we see that
the total DM emission follows the projected density profile over the
entire clusters. The LP model, however, give rise to a much flatter
profile in the central parts of the clusters. This is due to the IC
upscattered CMB photons that are upscattered on an electron and
positron population that suffers from efficent cooling on light from
SD. The IC emission from SD photons have a rather steep spectrum in
the center that follows the projected density profile which is
explained by the SD dominated cooling that cancels the spatial
dependece from the SD spectrum. In the left panel we also find some
variation in the profiles for the different clusters in the central
parts. Especially in the Fornax cluster, where we find a factor two
larger flux compared to the other clusters. This is due to more
efficient SD cooling of the dominating IC upscattered CMB photons in
Fornax; a low X-ray luminous cluster has a low energy density in
starlight and dust in our semi-analytic SD model. However, note that
the surface brightness from the SD component is suppresses at the
relative low energy of 1~GeV.  Since there is no enhancement from the
substructures, the overall mass normalization of the LP DM model has a
marginally negative trend $\sim\mvir^{-0.06}$. Hence, we only see a
very small difference in outer parts of the different clusters from
these type of models. The BM DM models have a normalization of
$\sim\mvir^{0.27}$, hence more massive clusters show a slightly higher
surface brightness. As also noticed in previous figure, the flux from
DM without substructures is dominated by the CRp induced emission for
all clusters.

>>>>

In Fig.~\ref{fig:radial_emis} we show the radius with the maximum
luminosity for substructures and smooth density profiles, both for
three different mass scales. We find that the smooth luminosity per
logarithmic radial bin peaks at $x \dd^2 L/\dd x^2+\dd L/\dd x == 0
\rightarrow r \propto \rvir/3c$. Hence the main contribution to the
luminosity for large mass scales, which are usually associated with a
low concentrated DM halos, is focused to the regime around 10\% of
$\rvir$.  It clearly shows how that the main contribution from the
substructures is coming from outer parts of the DM halos, while the
main contribution from the smooth density profiles peaks around the
scale radius. Not that, even though the bulk of substructures have
been erased in the central regions of DM halos, a cluster in
projection has a significant enhancement due to substructures at a
radius of just a few percent of $\rvir$.
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Luminosity dominated radius for different mass
    scales. The solid lines show the emissivity from a cuspy
    $\beta=1.0$ NFW density profile for three different mass scales;
    an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the emissivity from substructures for the same three mass
    scales. All emissivities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$.}
  \label{fig:radial_emis}
\end{figure}

@@

There is also a possibility to search for dark matter annihilation
indirectly in antimatter channels, like antiprotons or positrons. Here
diffusion is, however, very efficient which means that the directional
signature which is so important for gamma-rays, is completely
erased. This method has, however, recently been much in focus due to
the surprising rise with energy of the electron to proton ratio
measured by the PAMELA satellite experiment \cite{Adriani:2008zr} and
the ``bump'' in the total electron plus positron spectrum detected by
Fermi \cite{Abdo:2009zk}. The viable particle physics models
describing these results contain some non-trivial, or even unnatural
elements, however. First of all, annihilation to quarks has to be very
much suppressed, due to the non-observation of an enhanced antiproton
flux by PAMELA \cite{Adriani:2010rc}. This means that only so-called
{\em leptophilic} models, where the annihilation only goes to leptons
have to be considered. Here, one has to have a sizeable fraction going
to muons and tau leptons, as the shape of the spectrum disfavours
direct annihilation into electrons and positrons. To get agreement
with the measured absolute rates, one also has to have a large boost
factor (of the order of at least several hundred), which is difficult
to achieve with substructure alone. An additional enhancement of the
annihilation rate has to be invoked, for example given by the {\em
  Sommerfeld effect}. This basically means that the usual connection
between annihilation rate at decoupling (when relative velocities $v$
were of order $0.2 - 0.3$), which gives the relic density, and
present-day annihilation in the halo (with velocities $v\sim 10^{-3}$)
is broken. This would come about because the dark matter sector
contains additional light particles which mediates an interaction with
almost bound states which enhance the wave function at the relative
origin, and therefore the annihilation rate (for a pedagogical review
of viable scenarios, see \cite{ArkaniHamed:2008qn}). As a final
difficulty of these models, the lack of an inverse-Compton signature
from the galactic center means that only shallow density profiles can
describe the galactic dark matter density distribution
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the dark matter modelling of these effects, see
\cite{Bergstrom:2009ib}. For a recent treatment, showing still viable
models, see \cite{Finkbeiner:2010sm}.)

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, dark matter would
more likely be explained by the standard scenario of a WIMP, for
example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may gave interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical dark matter community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
dark matter halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few
hundred to a thousand in the solar neighborhood, as substructure
according to numerical simulations is most important in the outer
portions of the halo, due to tidal stripping of subhaloes in the inner
part. Another potentially very important effect which may explain the
large boost had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld may years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of dark matter
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}|,
$$ with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the
relative velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small
velocities. For a Yukawa-like particle of mass $m_\phi$, mediating a
weak attractive force with coupling constant $\alpha_Y$ between dark
matter particles $\chi$ with mass $m_\chi$, the small-velocity limit
of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the dark matter sector, one has to solve
numerically a coupled system of differential equations using
appropriate boundary conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters of
the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such a a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

@@

\subsubsection{Supersymmetric dark matter}
The most studied models for dark matter are supersymmetric ones,
especially models where the lightest supersymmetric particle, in most
models the lightest neutralino, is stable. The stability is assured in
viable SUSY models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for dark matter. In
addition, as the coupling to ordinary matter is given by gauge
couplings, the neutralino is automatically a WIMP
candidate. Unfortunately, the breaking of supersymmetry (which has to
be there since otherwise, e.g., there would exist a scalar electron
with the same mass as the electron) means that a large number of
essentially free parameters enter actual calculations.  The neutralino
is then a linear combination of the supersymmetric partners of the
U(1) gauge boson, one of the SU(2) gauge bosons, and the two Higgs
doublets in the Minimal Supersymmetric extension of the Standard Model
(the MSSM) with mixing coefficients which depend on the SUSY breaking
parameters \cite{1984NuPhB.238..453E}. The parameter space is very
large, too large in general to scan efficiently even with the most
powerful computers. Therefore, often simplifying assumptions are made
to limit the number of free parameters. Even then, a large number of
viable models is found which give a relic density consistent with the
WMAP data. This means that they are indeed very good templates for
WIMP dark matter, something that explains their popularity besides the
subjective statement that few more plausible models for dark matter
have been put forward so far.

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/lum.stars.eps}
\caption{\it accumulative luminosity}
 \label{fig:SD_lum}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/fit.porter.v2.eps}
\caption{\it Spectral dependence of radiation fields in a cluster of
  galaxies. The black dotted line in the left peak show the spectrum
  of CMB photons using a black body with a temperature of
  $2.73$~K. The crosses in the middle and right peaks represent the
  measured spectra from stars and dust (SD), respectively, and are
  derived in \cite{2006ApJ...648L..29P} for a galaxy. We normalize the
  individual SD spectrum separately using the observed luminosity from
  SD in clusters. The SD luminosity is related to the cluster mass
  through Eq, where we use have used the mass
  $\mvir=4.0\times10^{13}\msun$ in this figure. Finally we renormalize
  the SD spectra to the radius $r=0.071\rvir$, where the smooth energy
  density of the SD light (see Fig.~\ref{fig:SD_Edens}) equals the
  energy density of the CMB. The red dashed lines show the fitted SD
  spectral model.}
 \label{fig:SD_spectra}
\end{figure}
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/ucool.eps}
 \caption{\it Spatial dependence of the energy density of radiation
   fields in a cluster of galaxies. The energy density of the CMB,
   shown by the blue dotted line, is isotropic throughout the cluster,
   hence represented by a flat profile with
   $u_\rmn{cmb}=0.26\,\rmn{eV}\,\rmn{cm}^{-3}$. The energy density of
   the light from stars and dust (SD) is denoted by the red dashed
   line and the solid purple line for the total SD light and the
   smooth SD light, respectively. The SD light has been renormalized
   to a cluster with the mass $\mvir=4.0\times10^{13}\msun$. Finally we
   show the energy density two magnetic field models with a central
   magnetic field of 3~$\mu G$. The magnetic field scales with the gas
   density to the power $\alpha_\rmn{B}$; green dash-dotted line
   ($\alpha_\rmn{B}=0.5$) and orange dash-dotted line
   ($\alpha_\rmn{B}=0.7$). Note that the SD radiation is dominating
   the energy density inside $\sim0.1\,\rvir$.}
 \label{fig:SD_Edens}
\end{figure}

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by well
studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the angular extended Virgo cluster. It is quite striking how
constraining the upper limits for the Fornax cluster is compared to
the other clusters, especially in the $1-10$~GeV energy regime where
Fermi-LAT is most sensitive due to the maximized effective area. The
upper limits for Virgo and Perseus are background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. One should also note that the high CRp
induced flux from both Coma and Perseus is of the same magnitude as
the DM flux from LP models, making it difficult for indirect DM
searches in those clusters. Conversely is the Fornax cluster a great
target for indirect DM studies because the relative low gamma-ray flux
from CRps, absents of an active AGN, and high DM gamma-ray
flux. Actually, the Fermi-LAT 3 year data will be able to constrain
the smallest mass of substructures using the DM BM models. We have
also learned that the closest clusters are superior for DM searches
compared to further away clusters, even though the close by clusters
have a mass that is smaller. In fact, we can understand this by
looking at the luminosity-to-mass scaling relations. The gamma-ray
luminosity from the smooth density profile is given by
\begin{equation}
L_{\gamma,\rmn{sm}} \propto \int \dd V \rho(r)^2 \propto \frac{c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity including boost factors for the LP and
the BM models are given by
\begin{eqnarray}
\label{eq:LP_scaling}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \mvir^{0.73}\qquad
\,\,\,\,\,\,\rmn{and}\\
\label{eq:BM_scaling}
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \mvir^{1.06}\,,\qquad
\rmn{where} \\
B_\rmn{LP}&=&B_\sfe B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
B_\rmn{BM}&=&B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the mass with a factor two doubles the luminosity,
but increasing the distance by a factor two decrease the flux with a
factor four. It is also useful to relate the luminosities to a more
model independent variable as the maximum circular velocity given by
\begin{eqnarray}
V_\rmn{max} &=& V_\rmn{circ}(r_\rmn{max})\qquad\,\,\,\,\,\,\,\,\,\rmn{where}\nonumber\\
V_\rmn{circ}(r)&=& \sqrt{\frac{G_\rmn{N}\,M(<r)}{r}}\qquad\rmn{and}\nonumber\\
\frac{\dd V_\rmn{circ}(r)}{\dd r} &=& 0\quad \Longrightarrow\quad r_\rmn{max}\,.
\end{eqnarray}
We find that $V_\rmn{max}=650\,\rmn{km/s}$ of a large $10^{15}\,\msun$
cluster, and $V_\rmn{max}=420\,\rmn{km/s}$ of a small
$4\times10^{13}\,\msun$ group.

@@

NEED TO REWRITE, JUMPING BACK AND FORTH BETWEEN PANELS In order to
learn about the spatial distribution of the DM flux in various
clusters for different DM models and associated gamma-ray components,
we show in the left panel of Fig.~\ref{fig:SB_clu_nosub} the
brightness profiles for the smooth DM distribution for the same four
representative clusters that was shown previously in
Fig.~\ref{fig:SB_clu}. In addition, we show the spatial dependence of
the individual gamma-ray components for the Fornax cluster in the
right panel. We find that the spatial distribution of the DM emission
at 1~GeV in the outer parts of the all clusters have the same
shape. The shape trace the underlying density slope of $-3$, which is
expected for IC upscattered CMB photons. This picture is confirmed by
the components study in the right panel that show the neglible
contribution from the IC upscattered light from SD at this
energy. Since the surface brightness from BM models are dominated by
the continuum emission at each radial bin in all clusters, we see that
the total DM emission follows the projected density profile over the
entire clusters. The LP model, however, give rise to a much flatter
profile in the central parts of the clusters. This is due to the IC
upscattered CMB photons that are upscattered on an electron and
positron population that suffers from efficient cooling on light from
SD. The IC emission from SD photons have a rather steep spectrum in
the center that follows the projected density profile which is
explained by the SD dominated cooling that cancels the spatial
dependence from the SD spectrum. In the left panel we also find some
variation in the profiles for the different clusters in the central
parts. Especially in the Fornax cluster, where we find a factor two
larger flux compared to the other clusters. This is due to more
efficient SD cooling of the dominating IC upscattered CMB photons in
Fornax; a low X-ray luminous cluster has a low energy density in
starlight and dust in our semi-analytic SD model. However, note that
the surface brightness from the SD component is suppresses at the
relative low energy of 1~GeV.  Since there is no enhancement from the
substructures, the overall mass normalization of the LP DM model has a
marginally negative trend $\sim\mvir^{-0.06}$. Hence, we only see a
very small difference in outer parts of the different clusters from
these type of models. The BM DM models have a normalization of
$\sim\mvir^{0.27}$, hence more massive clusters show a slightly higher
surface brightness. As also noticed in previous figure, the flux from
DM without substructures is dominated by the CRp induced emission for
all clusters.
